On joint estimation of ancestral states and phylogenetic model parameters

### 2017-10-30 joint_inf.tex
addressing some comments

\section{Inconsistency of joint maximum likelihood}

@@

\section{Branch lengths}

We will now consider the problem of branch length estimation on the correct tree using joint ancestral state and branch length estimation.
We will consider two settings on the Farris topology, corresponding to whether some branch lengths are known ahead of time or not.
As above, we will assume that data is generated with two sister branches of fidelity $x$, with all other branches of length $y$.
%EM We will show XXX... and introduce \hat{w'} here.

\subsection{Fixed $x'$ and $y'$}

First, assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood for $(x',y',w')$.
%EM You are talking about a log likelihood here, right? So a likelihood being proportional to means that there is some additive constant. Makes things a little confusing.
If $x'$ and $y'$ are fixed, by collapsing like terms and ignoring terms with only $x'$ and $y'$, the likelihood in \eqref{eq:farris_likelihood} as a function of $w'$ is proportional to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
    &\qquad + p_{13}\cdot\log(1-w').
\end{align*}
Similarly, the other two possibilities for the likelihood are both proportional to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log(1+w').
\end{align*}
Since
$$
\log(1+w') \geq (1 - p_{13})\cdot\log(1+w') + p_{13}\cdot\log(1-w')
$$
with equality only when $w'=0$, the second likelihood is the likelihood with the maximum-likelihood ancestral state, and so we need only analyze that case.

@@

As a first attempt, simplify further to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2) \\
    &\qquad + (p_{1} + p_{3} + p_{12} + p_{23})          \cdot\log(1-x'^2) \\
    &\qquad + (p_{1} + p_{3})          \cdot\log(1+y'^2) \\
    &\qquad + (p_{2} + p_{123} + p_{12} + p_{23})          \cdot\log(1-y'^2) \\
    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
    &\qquad + \log(1+w')
\end{align*}
and
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2) \\
    &\qquad + (p_{1} + p_{3})          \cdot\log(1+y'^2) \\
    &\qquad + 2(1-p_{13})\log(1+x') \\
    &\qquad + 2(p_{1}+p_{3}+p_{12}+p_{23}+p_{13})\log(1-x') \\
    &\qquad + 2\log(1+y') \\
    &\qquad + 2(p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
    &\qquad + \log(1+w').
\end{align*}
Say we had estimated $\hat{w}'=1$.
Then we need is the region where $\hat{y}'>\hat{x}'$ and $0<\hat{x}',\hat{y}'<1$---the case for $\hat{x}'>\hat{y}'$ will be a reflection of this case.
Let's consider first $x'=y'$; if we can show that for $x,y<\epsilon$ where it looks like $\epsilon\approx 0.4$ then $\hat{w}'=1$ we are at least getting close to something.
So fix $x'=y'=u$ and get
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, u; \fullAncestralSplitPartitions(\tau_1,u))
    &\propto   p_{\emptyset}  \cdot\log(1+6u^2+u^4) \\
    &\qquad + p_{13}         \cdot\log(1-2u^2+u^4) \\
    &\qquad + (p_{1} + p_{2} + p_{3} + p_{123})          \cdot\log(1+u^2) \\
    &\qquad + (4-2p_{13})\log(1+u) \\
    &\qquad + 2(p_{1}+p_{2}+p_{3}+p_{123}+p_{13}+2p_{12}+2p_{23})\log(1-u) \\
\end{align*}
which is
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, u; \fullAncestralSplitPartitions(\tau_1,u))
    &\propto   p_{\emptyset}  \cdot\log(1+6u^2+u^4) \\
    &\qquad + (p_{13} + p_{1} + p_{2} + p_{3} + p_{123})          \cdot\log(1+u^2) \\
    &\qquad + (4-p_{13})\log(1+u) \\
    &\qquad + (2p_{1}+2p_{2}+2p_{3}+2p_{123}+3p_{13}+4p_{12}+4p_{23})\log(1-u). \\
\end{align*}

>>>>

\section{Inconsistency of joint maximum likelihood in estimating topology}

@@

\section{Inconsistency of joint maximum likelihood in estimating branch fidelities}

We will now consider the problem of branch fidelity estimation on the correct tree using joint ancestral state and branch fidelity estimation.
We will consider two settings on the Farris topology, corresponding to whether some branch fidelities are known ahead of time or not.
As above, we will assume that data is generated with two sister branches of fidelity $x$, with all other branches of length $y$.
We will show that, for a nontrivial subset of possible values for $x$ and $y$, the interior branch fidelity parameter $w'$ will be consistently overestimated as exactly equal to one instead of its true value of $y$.
That is, for the restricted case, with
$$
\hat{w}' = \arg\max_{w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x,y,w'\}; \fullAncestralSplitPartitions(\tau_1,\{x,y,w'\})),
$$
we show there are values for $x$ and $y$ where $\hat{w}'\equiv 1$.
In the general case, we do the same but with
$$
(\hat{x}', \hat{y}', \hat{w}') = \arg\max_{x',y',w'} \ \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})).
$$
%EM We will show XXX... and introduce \hat{w'} here.
%das: updated

\subsection{Restricted case: fixed $x'$ and $y'$}

First, assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood for $(x',y',w')$.
%EM You are talking about a log likelihood here, right? So a likelihood being proportional to means that there is some additive constant. Makes things a little confusing.
%das: updated
If $x'$ and $y'$ are fixed, by collapsing like terms and ignoring terms with only $x'$ and $y'$, the log-likelihood in \eqref{eq:farris_likelihood} as a function of $w'$ is
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\sim     p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
    &\qquad + p_{13}\cdot\log(1-w')
\end{align*}
where $\sim$ denotes equality up to an additive constant.
Similarly, the other two possibilities for the likelihood both satisfy
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\sim     p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log(1+w').
\end{align*}
Since
$$
\log(1+w') \geq (1 - p_{13})\cdot\log(1+w') + p_{13}\cdot\log(1-w')
$$
with equality only when $w'=0$, the second likelihood is the likelihood with the maximum-likelihood ancestral state, and so we need only analyze that case.

@@

As a first attempt, simplify further to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2) \\
    &\qquad + (p_{1} + p_{3} + p_{12} + p_{23})          \cdot\log(1-x'^2) \\
    &\qquad + (p_{1} + p_{3})          \cdot\log(1+y'^2) \\
    &\qquad + (p_{2} + p_{123} + p_{12} + p_{23})          \cdot\log(1-y'^2) \\
    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23}-2p_{13})\log(1+x') \\
    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23}+2p_{13})\log(1-x') \\
    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\log(1+y') \\
    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
    &\qquad + \log(1+w')
\end{align*}
and
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + (p_{2} + p_{123})          \cdot\log(1+x'^2) \\
    &\qquad + (p_{1} + p_{3})          \cdot\log(1+y'^2) \\
    &\qquad + 2(1-p_{13})\log(1+x') \\
    &\qquad + 2(p_{1}+p_{3}+p_{12}+p_{23}+p_{13})\log(1-x') \\
    &\qquad + 2\log(1+y') \\
    &\qquad + 2(p_{2}+p_{12}+p_{23}+p_{123})\log(1-y') \\
    &\qquad + \log(1+w').
\end{align*}
Say we had estimated $\hat{w}'=1$.
Then we need is the region where $\hat{y}'>\hat{x}'$ and $0<\hat{x}',\hat{y}'<1$---the case for $\hat{x}'>\hat{y}'$ will be a reflection of this case.
Let's consider first $x'=y'$; if we can show that for $x,y<\epsilon$ where it looks like $\epsilon\approx 0.4$ then $\hat{w}'=1$ we are at least getting close to something.
So fix $x'=y'=u$ and get
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, u; \fullAncestralSplitPartitions(\tau_1,u))
    &\sim     p_{\emptyset}  \cdot\log(1+6u^2+u^4) \\
    &\qquad + p_{13}         \cdot\log(1-2u^2+u^4) \\
    &\qquad + (p_{1} + p_{2} + p_{3} + p_{123})          \cdot\log(1+u^2) \\
    &\qquad + (4-2p_{13})\log(1+u) \\
    &\qquad + 2(p_{1}+p_{2}+p_{3}+p_{123}+p_{13}+2p_{12}+2p_{23})\log(1-u) \\
\end{align*}
which is
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, u; \fullAncestralSplitPartitions(\tau_1,u))
    &\sim     p_{\emptyset}  \cdot\log(1+6u^2+u^4) \\
    &\qquad + (p_{13} + p_{1} + p_{2} + p_{3} + p_{123})          \cdot\log(1+u^2) \\
    &\qquad + (4-p_{13})\log(1+u) \\
    &\qquad + (2p_{1}+2p_{2}+2p_{3}+2p_{123}+3p_{13}+4p_{12}+4p_{23})\log(1-u). \\
\end{align*}
